\begin{questions}
\question Write out all functions $f: \{1,2,3\}$ to $\{a,b\}$.  How many are there?  How many are injective?  How many are surjective?  How many are both?

	\begin{answer}
	There are 8 different functions.  For example, $f(1) = a$, $f(2) = a$, $f(3) = a$; or $f(1) = a$, $f(2) = b$, $f(3) = a$, and so on.  None of the functions are injective.  Exactly 6 of the functions are surjective.  No functions are both (since no functions here are injective).
	\end{answer}
	
	
	

\question Write out all functions $f: \{1,2\}$ to $\{a,b,c\}$.  How many are there?  How many are injective?  How many are surjective?  How many are both?

	\begin{answer}
	There are nine functions - you have a choice of three outputs for $f(1)$, and for each, you have three choices for the output $f(2)$.  Of these functions, 6 are injective, 0 are surjective, and 0 are both.
	\end{answer}
	
	

	


\question Consider functions $f: \{1,2,3,4\} \to \{a,b,c,d,e,f\}$.
\begin{parts}
  \part How many functions are there total?
  \part How many functions are injective?
  \part How many functions are surjective?
  \part How many functions have the property that $f(1) \ne a$ or $f(2) \ne b$, or both?
\end{parts}

	\begin{answer}
	\begin{parts}
	  \part $6^4 = 1296$, since there are six choices of where to send each of the 4 elements of the domain. %How many functions are there total?
	  \part $P(6, 4) = 6 \cdot 5 \cdot 4 \cdot 3 = 360$, since outputs cannot be repeated.  %How many functions are injective?
	  \part None. %How many functions are surjective?
	  \part There are $5 \cdot 6^3$ functions for which $f(1) \ne a$ and another $5 \cdot 6^3$ functions for which $f(2) \ne b$.  There are $5^2 \cdot 6^2$ functions for which both $f(1) \ne a$ and $f(2) \ne b$.  So the total number of functions for which $f(1) \ne a$ or $f(2) \ne b$ or both is
	  \[5 \cdot 6^3 + 5 \cdot 6^3 - 5^2 \cdot 6^2 = 1260\] %How many functions have the property that $f(1) \ne a$ or $f(2) \ne b$, or both?
	\end{parts}
	\end{answer}
	
	

\question Consider sets $A$ and $B$ with $|A| = 10$ and $|B| = 17$.
\begin{parts}
  \part How many functions $f: A \to B$ are there?
  \part How many functions $f: A \to B$ are injective?
\end{parts}

	\begin{answer}
	\begin{parts}
	  \part $17^{10}$ %How many functions $f: A \to B$ are there?
	  \part $P(17, 10)$  %How many functions $f: A \to B$ are injective?
	\end{parts}
	\end{answer}
	
	

\question Consider sets $A$ and $B$ with $|A| = 10$ and $|B| = 5$.  How many functions $f: A \to B$ are surjective?

	\begin{answer}
	$5^{10} - \left[{5 \choose 1}4^{10} - {5 \choose 2}3^{10} + {5 \choose 3}2^{10} - {5 \choose 4}1^{10}\right]$ %Consider sets $A$ and $B$ with $|A| = 10$ and $|B| = 5$.  How many functions $f: A \to B$ are surjective?
	\end{answer}
	
	


\question Let $A = \{1,2,3,4,5\}$.  How many injective functions $f:A \to A$ have the property that for each $x \in A$, $f(x) \ne x$?

	\begin{answer}
	$5! - \left[{5 \choose 1}4! - {5 \choose 2}3! + {5 \choose 3}2! - {5 \choose 4}1! + {5 \choose 5}0!\right]$.  This is a sneaky way to as for the number of derangements on 5 elements. %Let $A = \{1,2,3,4,5\}$.  How many injective functions $f:A \to A$ have the property that for each $x \in A$, $f(x) \ne x$?
	\end{answer}

	

\end{questions}