\begin{questions}



\question A \emph{multiset} is a collection of objects, just like a set, but can contain an object more than once (the order of the elements still doesn't matter though).  For example, $\{1,1, 2, 5, 5, 7\}$ is a multiset of size 6.
\begin{parts}
	\part How many sets of size 5 can be made using the 10 numeric digits 0 through 9?
	\part How many \emph{multi}sets of size 5 can be made using the 10 numeric digits 0 through 9?
\end{parts}

	\begin{answer}
		\begin{parts}
			\part ${10\choose 5}$.  We must select 5 of the 10 digits to put in the set.
			\part Use stars and bars: each star represents one of the 5 elements of the set, each bar represents a switch between digits.  So there are 5 stars and 9 bars, giving us ${14 \choose 9}$.
		\end{parts}
	\end{answer}



\question Each of the counting problems below can be solved with stars and bars.  For each, say what outcome the diagram
\[***|*||**|\]
represents, if there are the correct number of stars and bars for the problem.  Otherwise, say why the diagram does not represent any outcome, and what a correct diagram would look like.

\begin{parts}
	\part How many ways are there to select a handful of 6 jellybeans from a jar that contains 5 different flavors?
	\part How many ways can you distribute 5 identical lollipops to 6 kids?
	\part How many 6-letter words can you make using the 5 vowels?
	\part How many solutions are there to the equation $x_1 + x_2 + x_3 + x_4 = 6$
\end{parts}

	\begin{answer}
		\begin{parts}
			\part You take 3 strawberry, 1 lime, 0 licorice, 2 blueberry and 0 bubblegum.
			\part This is backwards.  We don't want the stars to represent the kids because the kids are not identical, but the stars are.  Instead we should use 5 stars (for the lollipops) and use 5 bars to switch between the 6 kids.  For example, **||***||| would represent the outcome with the first kid getting 2 lollipops, the third kid getting 3, and the rest of the kids getting none.
			\part This is the word AAAEOO.
			\part This doesn't represent a solution.  Each star should represent one of the 6 units that add up to 6, and the bars should \emph{switch} between the different variables.  We have one too many bars.
		\end{parts}
	\end{answer}



\question After gym class you are tasked with putting the 14 identical dodge-balls away into 5 bins.  
\begin{parts}
  \part How many ways can you do this if there are no restrictions?
  \part How many ways can you do this if each bin must contain at least one dodge-ball?
%  \part How many ways can you do this if no bin can hold more than 6 balls?
\end{parts}

	\begin{answer}
	 \begin{parts}
	   \part ${18 \choose 4}$.  Each outcome can be represented by a sequence of 14 stars and 4 bars. %How many ways can you do this if there are no restrictions?
	   \part ${13 \choose 4}$.  First put one ball in each bin.  This leaves 9 stars and 4 bars.%How many ways can you do this if each bin must contain at least one dodge-ball?
%	   \part ${18 \choose 4} - \left[ {5 \choose 1}{11 \choose 4} - {5 \choose 2}{4 \choose 4}\right]$.  Subtract all the distributions for which one or more bins contain 7 or more balls.  %How many ways can you do this if no bin can hold more than 6 balls?
	 \end{parts}
	\end{answer}
	
	


\question How many integer solutions are there to the equation $x + y + z = 8$
for which
\begin{parts}
  \part $x$, $y$, and $z$ are all positive?
  \part $x$, $y$, and $z$ are all non-negative?
  \part $x$, $y$, and $z$ are all greater than $-3$.
\end{parts}

	\begin{answer}
	\begin{parts}
	  \part ${7 \choose 2}$.  After each variable gets 1 star for free, we are left with 5 stars and 2 bars.  %$x$, $y$, and $z$ are all positive?
	  \part ${10 \choose 2}$.  We have 8 stars and 2 bars.  %$x$, $y$, and $z$ are all non-negative?
	  \part ${19 \choose 2}$.  This problem is equivalent to finding the number of solutions to $x' + y' + z' = 17$ where $x'$, $y'$ and $z'$ are non-negative.  (In fact, we really just do a substitution.  Let $x = x'- 3$, $y = y' - 3$ and $z = z' - 3$).  %$x$, $y$, and $z$ are all greater than $-3$.
	\end{parts}
	\end{answer}
	
	


\question Using the digits 2 through 8, find the number of different 5-digits numbers such that:

\begin{parts}
	\part Digits cannot be repeated and must be written in increasing order.  For example, 23678 is okay, but 32678 is not.
	\part Digits \emph{can} be repeated and must be written in non-decreasing order.  For example, 24448 is okay, but 24484 is not.
\end{parts}


	\begin{answer}
	\begin{parts}
		\part This is ${7 \choose 5}$.  We simply choose five of the seven digits and once chosen put them in increasing order.
		\part This requires stars and bars.  Use a star to represent each of the 5 digits in the number, and use their position relative to the bars to say what numeral fills that spot.  So we will have 5 stars and 6 bars, giving ${11 \choose 6}$.
	\end{parts}
	\end{answer}




\question When playing Yahtzee, you roll five regular 6-sided dice.  How many different outcomes are possible from a single roll?  The order of the dice does not matter.

	\begin{answer}
	${10 \choose 5}$.  We have 5 stars (the five dice) and 5 bars (the five switches between the number 1-6).
	\end{answer}
	
	

\question Your friend tells you she has 7 coins in her hand (just pennies, nickels, dimes and quarters).  If you guess how many of each kind of coin she has, she will give them to you.  If you guess randomly, what is the probability that you will be correct?

	\begin{answer}
	 We must figure out how many different combinations of 7 coins are possible.  Let a star represent each coin, and a bar represent switching between type of coin.  So for example **|*||**** represents 2 pennies, one nickel, no dimes and 4 quarters.  The number of such star and bar diagrams (with 7 stars and 3 bars) is ${10 \choose 3} = 120$.  Thus you have a 1 in 120 chance of guessing correctly.
	\end{answer}



\question How many integer solutions to $x_1 + x_2 + x_3 + x_4  = 25$ are there fore which $x_1 \ge 1$, $x_2 \ge 2$, $x_3 \ge 3$  and $x_4 \ge 4$?

	\begin{answer}
	${18 \choose 3}$.  Distribute 10 units to the variables before finding all solutions to $x_1' + x_2' + x_3' + x_4' = 15$ in non-negative integers.
	\end{answer}
	




\question Solve the three counting problems below.  Then say why it makes sense that they all have the same answer.  That is, say how you can interpret them as each other.

\begin{parts}
	\part How many ways are there to distribute 8 cookies to 3 kids?
	\part How many solutions in non-negative integers are there to the equation $x+y+z = 8$?
	\part How many different packs of 8 crayons can you make using crayons that come in red, blue and yellow?
\end{parts}



	\begin{answer}
	  The answer to each of the counting questions is ${10 \choose 2}$, using 8 stars and 2 bars.  The bars separate the kids, the variables, and the colors.  To see why these are really the same, notice that we can think of the 3 kids as named $x$, $y$, and $z$.  Or think of $x$ as red, $y$ as blue and $z$ as yellow.  Notice that in each case, the three things are distinguishable, while the 8 things (cookies, units, crayons) are all identical.
	\end{answer}


	

\question \textit{Conic}, your favorite math themed fast food drive-in offers 20 flavors which can be added to your soda.  You have enough money to buy a large soda with 4 added flavors.  How many different soda concoctions can you order if:
\begin{parts}
  \part you refuse to use any of the flavors more than once?
  \part you refuse repeats but care about the order the flavors are added?
  \part you allow yourself multiple shots of the same flavor?
  \part you allow yourself multiple shots, and care about the order the flavors are added?
\end{parts}

  \begin{answer}
   \begin{parts}
    \part ${20 \choose 4}$ (order does not matter and repeats are not allowed)
    \part $P(20, 4) = 20\cdot 19\cdot 18 \cdot 17$ (order matters and repeats are not allowed)
    \part ${23 \choose 19}$ (order does not matter and repeats are allowed; stars and bars)
    \part $20^4$ (order matters and repeats are allowed; 20 choices 4 times)
   \end{parts}
  \end{answer}
	
	
\end{questions}