\begin{questions}

	
\question Which (if any) of the graphs below are the same?  Which are different?  Explain.

\begin{center}
  \begin{tikzpicture}[yscale=.7]
    \draw[thick] (-2,0) \v -- (0,0) \v -- (2,0) \v -- (-1,2) \v -- (1,2) \v -- (0,0) -- (-1,2) (1,2) -- (-2,0);
  \end{tikzpicture}
  \hfill
  \begin{tikzpicture}[yscale=.7]
    \draw[thick] (-2,0) \v -- (0,0) \v -- (2,0) \v -- (0,1) \v -- (-2,0) -- (0,2) \v -- (2,0) (0,2) -- (0,1);
  \end{tikzpicture}
  \hfill
  \begin{tikzpicture}[yscale=1]
    \draw[thick] (-1, 0) \v -- (-1,1) \v -- (0,1) \v -- (1,1) \v -- (1,0) \v -- (0,1) -- (-1,0);
    \draw[thick] (-1,1) to [out=60, in=120] (1,1);
  \end{tikzpicture}
\end{center}

	\begin{answer}
		The first and the third graphs are the same, but the middle graph is different.
	\end{answer}
	
	
	


\question Which of the graphs in the previous question contain Euler paths or circuits?  Which of the graphs are planar?

	\begin{answer}
		The first (and third) graphs contain an Euler path.  All the graphs are planar.
	\end{answer}
	
	
	



\question Draw a graph which has an Euler circuit but is not planar.

	\begin{answer}
		For example, $K_5$.
	\end{answer}
	
	
	


\question Draw a graph which does not have an Euler path and is also not planar.

	\begin{answer}
		For example, $K_{3,3}$.
	\end{answer}
	
	
	
	

\question Suppose $G$ is a graph with $n$ vertices, each having degree 5. 
\begin{parts}
  \part For which values of $n$ does this make sense?
  \part For which values of $n$ does the graph have an Euler path?
  \part What is the smallest value of $n$ for which the graph might be planar? (tricky)
\end{parts}

	\begin{answer}
		\begin{parts}
		  \part Only if $n \ge 6$ and is even.%For which values of $n$ does this make sense?
		  \part None. %For which values of $n$ does the graph have an Euler path?
		  \part 12. Such a graph would have $\frac{5n}{2}$ edges.  If the graph is planar, then $n - \frac{5n}{2} + F = 2$ so there would be $\frac{4+3n}{2}$ faces.  Also, we must have $3F \le 2E$, since the graph is simple.  So we must have $3\frac{4 + 3n}{2} \le 5n$.  Solving for $n$ gives $n \ge 12$.%What is the smallest value of $n$ for which the graph might be planar? (tricky)
		\end{parts}
	\end{answer}
	
	
	

\end{questions}