
\begin{questions}
\question After gym class you are tasked with putting the 14 identical dodge-balls away into 5 bins.  
\begin{parts}
  \part How many ways can you do this if there are no restrictions?
  \part How many ways can you do this if each bin must contain at least one dodge-ball?
  \part How many ways can you do this if no bin can hold more than 6 balls?
\end{parts}

	\begin{answer}
	 \begin{parts}
	   \part ${18 \choose 4}$.  Each outcome can be represented by a sequence of 14 stars and 4 bars. %How many ways can you do this if there are no restrictions?
	   \part ${13 \choose 4}$.  First put one ball in each bin.  This leaves 9 stars and 4 bars.%How many ways can you do this if each bin must contain at least one dodge-ball?
	   \part ${18 \choose 4} - \left[ {5 \choose 1}{11 \choose 4} - {5 \choose 2}{4 \choose 4}\right]$.  Subtract all the distributions for which one or more bins contain 7 or more balls.  %How many ways can you do this if no bin can hold more than 6 balls?
	 \end{parts}
	\end{answer}
	
	


\question How many integer solutions are there to the equation $x + y + z = 8$
for which
\begin{parts}
  \part $x$, $y$, and $z$ are all positive?
  \part $x$, $y$, and $z$ are all non-negative?
  \part $x$, $y$, and $z$ are all greater than $-3$.
\end{parts}

	\begin{answer}
	\begin{parts}
	  \part ${7 \choose 2}$.  After each variable gets 1 star for free, we are left with 5 stars and 2 bars.  %$x$, $y$, and $z$ are all positive?
	  \part ${10 \choose 2}$.  We have 8 stars and 2 bars.  %$x$, $y$, and $z$ are all non-negative?
	  \part ${19 \choose 2}$.  This problem is equivalent to finding the number of solutions to $x' + y' + z' = 17$ where $x'$, $y'$ and $z'$ are non-negative.  (In fact, we really just do a substitution.  Let $x = x'- 3$, $y = y' - 3$ and $z = z' - 3$).  %$x$, $y$, and $z$ are all greater than $-3$.
	\end{parts}
	\end{answer}
	
	


\question When playing Yahtzee, you roll five regular 6-sided dice.  How many different outcomes are possible from a single roll?  The order of the dice does not matter.

	\begin{answer}
	${10 \choose 5}$.  We have 5 stars (the five dice) and 5 bars (the five switches between the number 1-6).
	\end{answer}
	
	


\question How many integer solutions to $x_1 + x_2 + x_3 + x_4  = 25$ are there fore which $x_1 \ge 1$, $x_2 \ge 2$, $x_3 \ge 3$  and $x_4 \ge 4$?

	\begin{answer}
	${18 \choose 3}$.  Distribute 10 units to the variables before finding all solutions to $x_1' + x_2' + x_3' + x_4' = 15$ in non-negative integers.
	\end{answer}
	
	


\question Write out all functions $f: \{1,2,3\}$ to $\{a,b\}$.  How many are there?  How many are injective?  How many are surjective?  How many are both?

	\begin{answer}
	There are 8 different functions.  For example, $f(1) = a$, $f(2) = a$, $f(3) = a$; or $f(1) = a$, $f(2) = b$, $f(3) = a$, and so on.  None of the functions are injective.  Exactly 6 of the functions are surjective.  No functions are both (since no functions here are injective).
	\end{answer}
	
	
	

\question Write out all functions $f: \{1,2\}$ to $\{a,b,c\}$.  How many are there?  How many are injective?  How many are surjective?  How many are both?

	\begin{answer}
	There are nine functions - you have a choice of three outputs for $f(1)$, and for each, you have three choices for the output $f(2)$.  Of these functions, 6 are injective, 0 are surjective, and 0 are both.
	\end{answer}
	
	
	

\question Consider the function $f:\{1,2,3,4,5\} \to \{1,2,3,4\}$ given by the table below:

\begin{center}
\begin{tabular}{c||c|c|c|c|c}
              $x$ & 1 & 2 & 3 & 4 & 5 \\ \hline
              $f(x)$ & 3 & 2 & 4 & 1 & 2
            \end{tabular}
\end{center}

\begin{parts}
  \part Is $f$ injective?  Explain.
  \part Is $f$ surjective?  Explain.
\end{parts}

	\begin{answer}
		\begin{parts}
		%Is $f$ injective?  Explain.
		\part $f$ is not injective, since $f(2) = f(5)$ - two different inputs have the same output. 
		% Is $f$ surjective?  Explain.
		\part $f$ is surjective, since every element of the codomain is an element of the range.
		\end{parts}
	\end{answer}
	
	
	

\question Consider the function $f:\{1,2,3,4\} \to \{1,2,3,4\}$ given by the graph below.
\begin{multicols}{2}
\begin{center}
  \begin{tikzpicture}[scale=1]
    \draw[thin, gray!50] (0,0) grid (4.5, 4.5);
    \draw[<->, thick] (0,4.5) node[left] {$f(x)$} -- (0,0) -- (4.5,0) node[below right] {$x$};
    \foreach \x in {1,2,3,4}
      \draw (\x,0) node[below] {\tiny \x} (0, \x) node[left] {\tiny \x};
    \fill (1,3) circle (.1) (2,4) circle (.1) (3,1) circle (.1) (4,3) circle (.1);
  \end{tikzpicture}
\end{center}

\begin{parts}
  \part Is $f$ injective?  Explain.
  \part Is $f$ surjective?  Explain.
\end{parts}
\end{multicols}


	\begin{answer}
		\begin{parts}
		%Is $f$ injective?  Explain.
		  \part $f$ is not injective, since $f(1) = 3$ and $f(4) = 3$.
		%   Is $f$ surjective?  Explain.
		  \part $f$ is not surjective, since there is no input which gives 2 as an output.
		\end{parts}
	\end{answer}
	
	
	


\question For each function given below, determine whether or not the function is injective and whether or not the function is surjective.
\begin{parts}
  \part $f:\N \to \N$ given by $f(n) = n+4$.
  \part $f:\Z \to \Z$ given by $f(n) = n+4$.
  \part $f:\Z \to \Z$ given by $f(n) = 5n - 8$.
  \part $f:\Z \to \Z$ given by $f(n) = \begin{cases}
                                         n/2 & \mbox{ if $n$ is even}\\
                                         (n+1)/2 & \mbox{ if $n$ is odd}.
                                       \end{cases}$
\end{parts}

	\begin{answer}
		\begin{parts}
		% $f:\N \to \N$ given by $f(n) = n+4$.
		  \part $f$ is injective, but not surjective.
		%   $f:\Z \to \Z$ given by $f(n) = n+4$.
		  \part $f$ is injective and surjective.
		  %$f:\Z \to \Z$ given by $f(n) = 5n - 8$.
		  \part $f$ is injective, but not surjective.
		%   $f:\Z \to \Z$ given by $f(n) = \begin{cases}
		%                                          n/2 & \mbox{ if $n$ is even}\\
		%                                          (n+1)/2 & \mbox{ if $n$ is odd}.
		%                                        \end{cases}$
		 \part $f$ is not injective, but is surjective.
		\end{parts}
	\end{answer}
	
	
	

%\question Let $A = \{1,2,3,\ldots,10\}$.  Consider the function $f:\pow(A) \to \N$ given by $f(B) = |B|$.  So $f$ takes a subset of $A$ as an input and outputs the cardinality of that set.  
%\begin{parts}
%  \part Is $f$ injective?  Prove your answer.
%  \part Is $f$ surjective?  Prove your answer.
%  \part Find $f\inv(1)$.
%  \part Find $f\inv(0)$.
%  \part Find $f\inv(12)$.
%\end{parts}
%
%	\begin{answer}
%		\begin{parts}
%		% Is $f$ injective?  Prove your answer.
%		  \part $f$ is not injective.  To prove this, we must simply find two different elements of the domain which map to the same element of the codomain.  Since $f(\{1\}) = 1$ and $f(\{2\}) = 1$, we see that $f$ is not injective.
%		%   Is $f$ surjective?  Prove your answer.
%		  \part $f$ is not surjective.  The largest subset of $A$ is $A$ itself, and $|A| = 10$.  So no natural number greater than 10 will ever be an output.
%		%   Find $f\inv(1)$.
%		  \part $f\inv(1) = \{\{1\}, \{2\}, \{3\}, \ldots \{10\}\}$ (the set of all the singleton subsets of $A$).
%		%   Find $f\inv(0)$.
%		  \part $f\inv(0) = \{\emptyset\}$.  Note, it would be wrong to write $f\inv(0) = \emptyset$ - that would claim that there is no input which has 0 as an output.
%		   % Find $f\inv(12)$.
%		  \part $f\inv(12) = \emptyset$, since there are no subsets of $A$ with cardinality 12.
%		\end{parts}
%	\end{answer}
	
	
	

%\question Let $A = \{n \in \N \st 0 \le n \le 999\}$ be the set of all numbers with three or fewer digits.  Define the function $f:A \to \N$ by $f(abc) = a+b+c$, where $a$, $b$, and $c$ are the digits of the number in $A$.  For example, $f(253) = 2 + 5 + 3 =  10$.
%\begin{parts}
%  \part Find $f\inv(3)$.
%  \part Find $f\inv(28)$.
%  \part Use one of the parts above to prove that $f$ is not injective.
%  \part Use one of the parts above to prove that $f$ is not surjective.
%\end{parts}
%
%	\begin{answer}
%		\begin{parts}
%		% Find $f\inv(3)$.
%		  \part $f\inv(3) = \{003, 030, 300, 012, 021, 102, 201, 120, 210, 111\}$
%		%   Find $f\inv(28)$.
%		  \part $f\inv(28) = \emptyset$ (since the largest sum of three digits is $9+9+9 = 27$)
%		%   Use one of the parts above to prove that $f$ is not injective.
%		  \part Part (a) proves that $f$ is not injective - the output 3 is assigned to 10 different inputs.
%		%   Use one of the parts above to prove that $f$ is not surjective.
%		  \part Part (b) proves that $f$ is not surjective - there is an element of the codomain (28) which is assigned to no inputs.
%		\end{parts}
%	\end{answer}
	
	
	

%\question Find a set $X$ and a function $f:X \to \N$ so that $f\inv(0) \cup f\inv(1) = X$.
%
%	\begin{answer}
%		$X$ can really be any set, as long as $f(x) = 0$ or $f(x) = 1$ for every $x \in X$.  For example, $X = \N$ and $f(n) = 0$ works.
%	\end{answer}
	
	
	

\question What can you deduce about the sets $X$ and $Y$ if you know,
\begin{parts}
  \part there is a injective function $f:X \to Y$.  Explain.
  \part there is a surjective function $f:X \to Y$.  Explain.
  \part there is a bijection $f:X \to Y$.  Explain.
\end{parts}

	\begin{answer}
		\begin{parts}
		% there is a injective function $f:X \to Y$.  Explain.
		  \part $|X| \le |Y|$ - otherwise two or more of the elements of $X$ would need to map to the same element of $Y$.
		%   there is a surjective function $f:X \to Y$.  Explain.
		  \part $|X| \ge |Y|$ - otherwise there would be one or more elements of $Y$ which were never an output.
		%   there is a bijection $f:X \to Y$.  Explain.
		  \part $|X| = |Y|$.  This is the only way for both of the above to occur.
		\end{parts}
	\end{answer}
	
	
	

\question Suppose $f:X \to Y$ is a function.  Which of the following are possible?  Explain.
\begin{parts}
  \part $f$ is injective but not surjective.
  \part $f$ is surjective but not injective.
  \part $|X| = |Y|$ and $f$ is injective but not surjective.
  \part $|X| = |Y|$ and $f$ is surjective but not injective.
  \part $|X| = |Y|$, $X$ and $Y$ are finite, and $f$ is injective but not surjective.
  \part $|X| = |Y|$, $X$ and $Y$ are finite, and $f$ is surjective but not injective.
\end{parts}

	\begin{answer}
		\begin{parts}
		% $f$ is injective but not surjective.
		  \part Yes. (Can you give an example?)
		  \part Yes. %$f$ is surjective but not injective.
		  \part Yes. %$|X| = |Y|$ and $f$ is injective but not surjective.
		  \part Yes. %$|X| = |Y|$ and $f$ is surjective but not injective.
		  \part No. %$|X| = |Y|$, $X$ and $Y$ are finite, and $f$ is injective but not surjective.
		  \part No. %$|X| = |Y|$, $X$ and $Y$ are finite, and $f$ is surjective but not injective.
		\end{parts}
	\end{answer}
	
	
	

%\question Consider the function $f:\Z \to \Z$ given by $f(n) = \begin{cases}
%                                                                 n+1 & \mbox{ if $n$ is even}\\
%                                                                 n-3 & \mbox{ if $n$ is odd}.
%                                                               \end{cases}$
%\begin{parts}
%  \part Is $f$ injective?  Prove your answer.
%  \part Is $f$ surjective?  Prove your answer.
%\end{parts}
%
%	\begin{answer}
%		\begin{parts}
%		% Is $f$ injective?  Prove your answer.
%		  \part $f$ is injective. 
%		  \begin{proof}
%		   Let $x$ and $y$ be elements of the domain $\Z$.  Assume $f(x) = f(y)$.  If $x$ and $y$ are both even, then $f(x) = x+1$ and $f(y) = y+1$.  Since $f(x) = f(y)$, we have $x + 1 = y + 1$ which implies that $x = y$.  Similarly, if $x$ and $y$ are both odd, then $x - 3 = y-3$ so again $x = y$.  The only other possibility is that $x$ is even an $y$ is odd (or visa-versa).  But then $x + 1$ would be odd and $y - 3$ would be even, so it cannot be that $f(x) = f(y)$.  Therefore if $f(x) = f(y)$ we then have $x = y$, which proves that $f$ is injective.
%		  \end{proof}
%		% Is $f$ surjective?  Prove your answer.
%		  \part $f$ is surjective.
%		  \begin{proof}
%		   Let $y$ be an element of the codomain $\Z$.  We will show there is an element $n$ of the domain ($\Z$) such that $f(n) = y$.  There are two cases.  First, if $y$ is even, then let $n = y+3$.  Since $y$ is even, $n$ is odd, so $f(n) = n-3 = y+3-3 = y$ as desired.  Second, if $y$ is odd, then let $n = y-1$.  Since $y$ is odd, $n$ is even, so $f(n) = n+1 = y-1+1 = y$ as needed.  Therefore $f$ is surjective.
%		  \end{proof}
%		
%		\end{parts}
%	\end{answer}
	
	
	


\question Consider functions $f: \{1,2,3,4\} \to \{a,b,c,d,e,f\}$.
\begin{parts}
  \part How many functions are there total?
  \part How many functions are injective?
  \part How many functions are surjective?
  \part How many functions have the property that $f(1) \ne a$ or $f(2) \ne b$, or both?
\end{parts}

	\begin{answer}
	\begin{parts}
	  \part $6^4 = 1296$, since there are six choices of where to send each of the 4 elements of the domain. %How many functions are there total?
	  \part $P(6, 4) = 6 \cdot 5 \cdot 4 \cdot 3 = 360$, since outputs cannot be repeated.  %How many functions are injective?
	  \part None. %How many functions are surjective?
	  \part There are $5 \cdot 6^3$ functions for which $f(1) \ne a$ and another $5 \cdot 6^3$ functions for which $f(2) \ne b$.  There are $5^2 \cdot 6^2$ functions for which both $f(1) \ne a$ and $f(2) \ne b$.  So the total number of functions for which $f(1) \ne a$ or $f(2) \ne b$ or both is
	  \[5 \cdot 6^3 + 5 \cdot 6^3 - 5^2 \cdot 6^2 = 1260\] %How many functions have the property that $f(1) \ne a$ or $f(2) \ne b$, or both?
	\end{parts}
	\end{answer}
	
	

\question Consider sets $A$ and $B$ with $|A| = 10$ and $|B| = 17$.
\begin{parts}
  \part How many functions $f: A \to B$ are there?
  \part How many functions $f: A \to B$ are injective?
\end{parts}

	\begin{answer}
	\begin{parts}
	  \part $17^{10}$ %How many functions $f: A \to B$ are there?
	  \part $P(17, 10)$  %How many functions $f: A \to B$ are injective?
	\end{parts}
	\end{answer}
	
	



\question Consider sets $A$ and $B$ with $|A| = 10$ and $|B| = 5$.  How many functions $f: A \to B$ are surjective?

	\begin{answer}
	$5^{10} - \left[{5 \choose 1}4^{10} - {5 \choose 2}3^{10} + {5 \choose 3}2^{10} - {5 \choose 4}1^{10}\right]$ %Consider sets $A$ and $B$ with $|A| = 10$ and $|B| = 5$.  How many functions $f: A \to B$ are surjective?
	\end{answer}
	
	


\question Let $A = \{1,2,3,4,5\}$.  How many injective functions $f:A \to A$ have the property that for each $x \in A$, $f(x) \ne x$?

	\begin{answer}
	$5! - \left[{5 \choose 1}4! - {5 \choose 2}3! + {5 \choose 3}2! - {5 \choose 4}1! + {5 \choose 5}0!\right]$.  This is a sneaky way to as for the number of derangements on 5 elements. %Let $A = \{1,2,3,4,5\}$.  How many injective functions $f:A \to A$ have the property that for each $x \in A$, $f(x) \ne x$?
	\end{answer}
	
	


\question Ten ladies of a certain age drop off their red hats at the hat check of a museum.  As they are leaving, the hat check attendant gives the hats back randomly.  In how many ways can exactly six of the ladies receive their own hat (and the other four not)?

	\begin{answer}
	${10 \choose 6}\left(4! - \left[{4 \choose 1} 3! - {4 \choose 2}2! + {4 \choose 3}1! - {4 \choose 4}0!\right]\right)$.  We choose 6 of the 10 ladies to get their own hat, and the multiply by the number of ways the remaining hats can be deranged. 
	\end{answer}
	
	



	


\question You have 9 presents to give to your 4 kids.  How many ways can this be done if
\begin{parts}
  \part The presents are identical, and each kid gets at least one present?
  \part The presents are identical, and some kids might get no presents?
  \part The presents are unique, and some kids might get no presents?
  \part the presents are unique and each kid gets at least one present?
\end{parts}

	\begin{answer}
	\begin{parts}
	  \part ${8 \choose 3}$, after giving one present to each kid, you are left with 5 presents (stars) which need to be divide among the 4 kids (giving 3 bars). %The presents are identical, and each kid gets at least one present?
	  \part ${12 \choose 3}$.  You have 9 stars and 3 bars.  %The presents are identical, and some kids might get no presents?
	  \part $4^9$.  You have 4 choices for whom to give each present.  This is like making a function from the set of present to the set of kids. %The presents are unique, and some kids might get no presents?
	  \part $4^9 - \left[{4 \choose 1}3^9 - {4\choose 2}2^9 + {4 \choose 3}1^9 \right]$.  Now the function from the set of present to the set of kids must be surjective. %the presents are unique and each kid gets at least one present?
	\end{parts}
	\end{answer}
	
	


\newpage

\question For each of the following counting problems, say whether the answer is


\begin{itemize}
\begin{multicols}{3}
  \item[A] ${10\choose 4}$
  \item[B] $P(10,4)$
  \item[C] Neither
  \end{multicols}
\end{itemize}

If you answer is ``Neither,'' say what the answer should be instead.
\begin{parts}
  \part How many shortest lattice paths are there from $(0,0)$ to $(10,4)$?
  \part If you have 10 bow ties, and you want to select 4 of them for next week, how many choices do you have?
  \part Suppose you have 10 bow ties and you will wear one on each of the next 4 days.  How many choices do you have?
  \part If you want to wear 4 of your 10 bow ties next week (Monday through Sunday), how many ways can this be accomplished?
  \part Out of a group of 10 classmates, how many ways can you rank your top 4 friends?
  \part If 10 students come to their professor's office but only 4 can fit at a time, how different combinations of 4 students can see the prof first?
  \part How many 4 letter words can be made from the first 10 letters of the alphabet?
  \part How many ways can you make the word ``cake'' from the first 10 letters of the alphabet?
  \part How many ways are there to distribute 10 apples among 4 children?
  \part If you have 10 kids (and live in a shoe) and 4 types of cereal, how many ways can your kids eat breakfast?
  \part How many ways can you arrange exactly 4 ones in a string of 10 binary digits?
  \part You want to select 4 single digit numbers as your lotto picks.  How many choices do you have?
  \part 10 kids want ice-cream.  You have 4 varieties.  How many ways are there to give the kids as much ice-cream as they want?
  \part How many 1-1 functions are there from $\{1,2,\ldots, 10\}$ to $\{a,b,c,d\}$?
  \part How many surjective functions are there from $\{1,2,\ldots, 10\}$ to $\{a,b,c,d\}$?
  \part Each of your 10 bow ties match 4 pairs of suspenders.  How many outfits can you make?
  \part After the party, the 10 kids each choose one of 4 party-favors.  How many outcomes?
  \part How many 6-elements subsets are there of the set $\{1,2,\ldots, 10\}$
  \part How many ways can you split up 11 kids into 5 teams?
  \part How many solutions are there to $x_1 + x_2 + \cdots + x_5 = 6$ where each $x_i$ is non-negative?
  \part Your band goes on tour.  There are 10 cities within driving distance, but only enough time to play 4 of them.  How many choices do you have for the cities on your tour?
  \part In how many different ways can you play the 4 cities you choose?
  \part Out of the 10 breakfast cereals available, you want to have 4 bowls.  How many ways can you do this?
  \part There are 10 types of cookies available.  You want to make a 4 cookie stack.  How many different stacks can you make?
  \part From you home at (0,0) you want to go to either the donut shop at (5,4) or the one at (3,6).  How many paths could you take?
  \part How many 10-digit numbers do not contain a sub-string of 4 repeated digits?
\end{parts}

	\begin{answer}
	\begin{parts}
	    \part Neither.  ${14 \choose 4}$.
	  \part ${10\choose 4}$
	  \part $P(10,4)$, since order is important.
	  \part Neither. Assuming you will wear each of the 4 ties on just 4 of the 7 days, without repeats: ${10\choose 4}P(7,4)$.
	  \part $P(10,4)$
	  \part ${10\choose 4}$
	  \part Neither. Since you could repeat letters: $10^4$. If no repeats are allowed, it would be $P(10,4)$.
	  \part Neither.  Actually, ``k'' is the 11th letter of the alphabet, so the answer is 0.  If ``k'' was among the first 10 letters, there would only be 1 way - write it down.
	  \part Neither.  Either ${9\choose 3}$ (if every kid gets an apple) or ${13 \choose 3}$ (if appleless kids are allowed).
	  \part Neither.  Note that this could not be ${10 \choose 4}$ since the 10 things and 4 things are from different groups.  $4^{10}$
	  \part ${10 \choose 4}$ - don't be fooled by the ``arrange'' in there - you are picking 4 out of 10 {\em spots} to put the 1's. 
	  \part ${10 \choose 4}$ (assuming order is irrelevant). 
	  \part Neither.  $16^{10}$ (each kid chooses yes or no to 4 varieties).
	  \part Neither.  0.
	  \part Neither.  $4^{10} - [{4\choose 1}3^{10} - {4\choose 2}2^{10} + {4 \choose 3}1^{10}]$
	  \part Neither.  $10\cdot 4$.
	  \part Neither. $4^{10}$.
	  \part ${10 \choose 4}$ (which is the same as ${10 \choose 6}$).
	  \part Neither.  If all the kids were identical, and you wanted no empty teams, it would be ${10 \choose 4}$.  Instead, this will be the same as the number of surjective functions from a set of size 11 to a set of size 5. 
	  \part ${10 \choose 4}$
	  \part ${10 \choose 4}$
	  \part Neither.  $4!$.
	  \part Neither.  It's ${10 \choose 4}$ if you won't repeat any choices.  If repetition is allowed, then this becomes $x_1 + x_2 + \cdots +x_{10} = 4$, which has ${13 \choose 9}$ solutions in non-negative integers.
	  \part Neither.  Since repetition of cookie type is allowed, the answer is $10^4$.  Without repetition, you would have $P(10,4)$.
	  \part ${10 \choose 4}$ since that is equal to ${9 \choose 4} + {9 \choose 3}$.
	  \part Neither.  It will be a complicated (possibly PIE) counting problem.
	\end{parts}
	\end{answer}
	
	


\end{questions}



