\begin{questions}

\question Find the next 2 terms in the sequence $3, 5, 11, 21, 43, 85\ldots.$.  Then give a recursive definition for the sequence.  Finally, use the characteristic root technique to find a closed formula for the sequence.

	\begin{answer}
		171 and 341.  $a_n = a_{n-1} + 2a_{n-2}$ with $a_0 = 3$ and $a_1 = 5$.  Closed formula: $a_n = \frac{8}{3}2^n + \frac{1}{3}(-1)^n$
	\end{answer}
	
	
	


\question Solve the recurrence relation $a_n = a_{n-1} + 2^n$ with $a_0 = 5$.

	\begin{answer}
		By telescoping or iteration.  $a_n = 3 + 2^{n+1}$
	\end{answer}
	
	
	


\question Show that $4^n$ is a solution to the recurrence relation $a_n = 3a_{n-1} + 4a_{n-2}$.

	\begin{answer}
		We claim $a_n = 4^n$ works.  Plug it in: $4^n = 3(4^{n-1}) + 4(4^{n-2})$.  This works - just simplify the right hand side.
	\end{answer}
	
	
	


\question Find the solution to the recurrence relation $a_n = 3a_{n-1} + 4a_{n-2}$ with initial terms $a_0 = 2$ and $a_1 = 3$.

	\begin{answer}
		By the Characteristic Root Technique.  $a_n = 4^n + (-1)^n$.
	\end{answer}
	
	
	


\question Find the solution to the recurrence relation $a_n = 3a_{n-1} + 4a_{n-2}$ with initial terms $a_0 = 5$ and $a_1 = 8$.

	\begin{answer}
		$a_n = \frac{13}{5} 4^n + \frac{12}{5} (-1)^n$
	\end{answer}
	
	
	


\question Solve the recurrence relation $a_n = 2a_{n-1} - a_{n-2}$.
\begin{parts}
  \part What is the solution if the initial terms are $a_0 = 1$ and $a_1 = 2$?
  \part What do the initial terms need to be in order for $a_9 = 30$?
  \part For which $x$ are there initial terms which make $a_9 = x$?
\end{parts}

	\begin{answer}
		The general solution is $a_n = a + bn$ where $a$ and $b$ depend on the initial conditions.  %Solve the recurrence relation $a_n = 2a_{n-1} - a_{n-2}$.
		\begin{parts}
		  \part $a_n = 1 + n$
		  %What is the solution if the initial terms are $a_0 = 1$ and $a_1 = 2$?
		  \part For example, we could have $a_0 = 21$ and $a_1 = 22$.  %What do the initial terms need to be in order for $a_9 = 30$?
		  \part For every $x$ - take $a_0 = x-9$ and $a_1 = x-8$.  %For which $x$ are there initial terms which make $a_9 = x$?
		\end{parts}
	\end{answer}
	
	
	


\question Solve the recurrence relation $a_n = 3a_{n-1} + 10a_{n-2}$ with initial terms $a_0 = 4$ and $a_1 = 1$.  

	\begin{answer}
		$a_n = \frac{19}{7}(-2)^n + \frac{9}{7}5^n$
		%Solve the recurrence relation $a_n = 3a_{n-1} + 10a_{n-2}$ with initial terms $a_0 = 4$ and $a_1 = 1$.  
	\end{answer}
	
	

\end{questions}