\documentclass[11pt]{exam}

\usepackage{amsmath, amssymb, amsthm, multicol}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{tikz}

\def\d{\displaystyle}
\def\b{\mathbf}
\def\R{\mathbb{R}}
\def\Z{\mathbb{Z}}
\def\N{\mathbb{N}}
\def\pow{\mathcal{P}}
\def\st{~:~}
\def\bar{\overline}
\def\inv{^{-1}}
\def\imp{\rightarrow}
\def\and{\wedge}


%\pointname{pts}
\pointsinmargin
\marginpointname{pts}
\addpoints
\pagestyle{head}
%\printanswers

\firstpageheader{Math 228}{\bf Practice Problems 9: Advanced Counting}{Spring 2012}


\begin{document}
\noindent \textbf{Instructions}: The problems below are purely for you to practice.  I will not collect these, but it is still a good idea to write out your solutions in full.  Any of these problems or problems similar are fair game for quizzes and exams.  

\begin{questions}
\question Consider functions $f: \{1,2,3,4\} \to \{a,b,c,d,e,f\}$.
\begin{parts}
  \part How many functions are there total?
  \part How many functions are one-to-one?
  \part How many functions are onto?
  \part How many functions have the property that $f(1) \ne a$ or $f(2) \ne b$, or both?
\end{parts}

\question Consider sets $A$ and $B$ with $|A| = 10$ and $|B| = 17$.
\begin{parts}
  \part How many functions $f: A \to B$ are there?
  \part How many functions $f: A \to B$ are one-to-one?
\end{parts}

\question Consider sets $A$ and $B$ with $|A| = 10$ and $|B| = 5$.  How many functions $f: A \to B$ are onto?

\question Let $A = \{1,2,3,4,5\}$.  How many one-to-one functions $f:A \to A$ have the property that for each $x \in A$, $f(x) \ne x$?

\question Ten ladies of a certain age drop off their red hats at the hat check of a museum.  As they are leaving, the hat check attendant gives the hats back randomly.  What is the probability that exactly six of the ladies receive their own hat (and the other four do not)?

\question After gym class you are tasked with putting the 14 identical dodge-balls away into 5 bins.  
\begin{parts}
  \part How many ways can you do this if there are no restrictions?
  \part How many ways can you do this if each bin must contain at least one dodge-ball?
  \part How many ways can you do this if no bin can hold more than 6 balls?
\end{parts}

\question How many integer solutions are there to the equation $x + y + z = 8$
for which
\begin{parts}
  \part $x$, $y$, and $z$ are all positive?
  \part $x$, $y$, and $z$ are all non-negative?
  \part $x$, $y$, and $z$ are all greater than $-3$.
\end{parts}

\question When playing Yahtzee, you roll five regular 6-sided dice.  How many different outcomes are possible from a single roll?  The order of the dice does not matter.

\question How many integer solutions to $x_1 + x_2 + x_3 + x_4  = 25$ are there fore which $x_1 \ge 1$, $x_2 \ge 2$, $x_3 \ge 3$  and $x_4 \ge 4$?

\question You have 9 presents to give to your 4 kids.  How many ways can this be done if
\begin{parts}
  \part The presents are identical, and each kid gets at least one present?
  \part The presents are identical, and some kids might get no presents?
  \part The presents are unique, and some kids might get no presents?
  \part the presents are unique and each kid gets at least one present?
\end{parts}
\newpage

\question For each of the following counting problems, say whether the answer is

\centerline{
\begin{oneparchoices}
  \choice ${10\choose 4}$
  \choice $P(10,4)$
  \choice Neither
\end{oneparchoices}
}
If you answer is ``Neither,'' say what the answer should be instead.
\begin{parts}
  \part How many shortest lattice paths are there from $(0,0)$ to $(10,4)$?
  \part If you have 10 bow ties, and you want to select 4 of them for next week, how many choices do you have?
  \part Suppose you have 10 bow ties and you will wear one on each of the next 4 days.  How many choices do you have?
  \part If you want to wear 4 of your 10 bow ties next week (Monday through Sunday), how many ways can this be accomplished?
  \part Out of a group of 10 classmates, how many ways can you rank your top 4 friends?
  \part If 10 students come to their professor's office but only 4 can fit at a time, how different combinations of 4 students can see the prof first?
  \part How many 4 letter words can be made from the first 10 letters of the alphabet?
  \part How many ways can you make the word ``cake'' from the first 10 letters of the alphabet?
  \part How many ways are there to distribute 10 apples among 4 children?
  \part If you have 10 kids (and live in a shoe) and 4 types of cereal, how many ways can your kids eat breakfast?
  \part How many ways can you arrange exactly 4 ones in a string of 10 binary digits?
  \part You want to select 4 single digit numbers as your lotto picks.  How many choices do you have?
  \part 10 kids want ice-cream.  You have 4 varieties.  How many ways are there to give the kids as much ice-cream as they want?
  \part How many 1-1 functions are there from $\{1,2,\ldots, 10\}$ to $\{a,b,c,d\}$?
  \part How many onto functions are there from $\{1,2,\ldots, 10\}$ to $\{a,b,c,d\}$?
  \part Each of your 10 bow ties match 4 pairs of suspenders.  How many outfits can you make?
  \part After the party, the 10 kids each choose one of 4 party-favors.  How many outcomes?
  \part How many 6-elements subsets are there of the set $\{1,2,\ldots, 10\}$
  \part How many ways can you split up 11 kids into 5 teams?
  \part How many solutions are there to $x_1 + x_2 + \cdots + x_5 = 6$ where each $x_i$ is non-negative?
  \part Your band goes on tour.  There are 10 cities within driving distance, but only enough time to play 4 of them.  How many choices do you have for the cities on your tour?
  \part In how many different ways can you play the 4 cities you choose?
  \part Out of the 10 breakfast cereals available, you want to have 4 bowls.  How many ways can you do this?
  \part There are 10 types of cookies available.  You want to make a 4 cookie stack.  How many different stacks can you make?
  \part From you home at (0,0) you want to go to either the donut shop at (5,4) or the one at (3,6).  How many paths could you take?
  \part How many 10-digit numbers do not contain a sub-string of 4 repeated digits?
\end{parts}
\end{questions}


\end{document}


