\documentclass[10pt]{exam}

\usepackage{amsmath, amssymb, amsthm, multicol}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{tikz}

\def\d{\displaystyle}
\def\b{\mathbf}
\def\R{\mathbb{R}}
\def\Z{\mathbb{Z}}
\def\N{\mathbb{N}}
\def\pow{\mathcal{P}}
\def\st{~:~}
\def\bar{\overline}
\def\inv{^{-1}}
\def\imp{\rightarrow}
\def\and{\wedge}


%\pointname{pts}
\pointsinmargin
\marginpointname{pts}
\addpoints
\pagestyle{head}
%\printanswers

\firstpageheader{Math 228}{\bf Practice Problems 8: Binomial Identities}{Spring 2012}


\begin{document}
\noindent \textbf{Instructions}: The problems below are purely for you to practice.  I will not collect these, but it is still a good idea to write out your solutions in full.  Any of these problems or problems similar are fair game for quizzes and exams.  

\begin{questions}
\question Give a combinatorial proof for the identity $1 + 2 + 3 + \cdots + n = {n+1 \choose 2}$. 

\question A woman is getting married.  She has 15 best friends but can only select 6 of them to be her bride's maids, one of which needs to be her maid of honor.  How many ways can she do this?
\begin{parts}
 \part What if she first selects the 6 bride's maids, and then selects one of them to be the maid of honor?
 \part What if she first selects her maid of honor, and then 5 other bride's maids?
 \part Explain why $6 {15 \choose 6} = 15 {14 \choose 5}$.
\end{parts}

\question Consider the bit strings in $\b B^6_2$ (bit strings of length 6 and weight 2).
\begin{parts}
 \part How many of those bit strings start with 1?
 \part How many of those bit strings start with 01?
 \part How many of those bit strings start with 001?
 \part Are there any other strings we have not counted yet?  Which ones, and how many are there?
 \part How many bit strings are there total in $\b B^6_2$?
 \part What binomial identity have you just given a combinatorial proof for?
\end{parts}


\question Let's count {\em ternary} digit strings - strings in which each digit can be 0, 1, or 2.
\begin{parts}
 \part How many ternary digit strings contain exactly $n$ digits?  
 \part How many ternary digit strings contain exactly $n$ digits and $n$ 2's.
 \part How many ternary digit strings contain exactly $n$ digits and $n-1$ 2's.  (Hint: where can you put the non-2 digit, and then what could it be?)
 \part How many ternary digit strings contain exactly $n$ digits and $n-2$ 2's.  (Hint: see previous hint)
 \part How many ternary digit strings contain exactly $n$ digits and $n-k$ 2's.
 \part How many ternary digit strings contain exactly $n$ digits and no 2's. (Hint: what kind of a string is this?)
 \part Use the above parts to give a combinatorial proof for the identity
 \[{n \choose 0} + 2{n \choose 1} + 2^2{n \choose 2} + 2^3{n \choose 3} + \cdots + 2^n{n \choose n} = 3^n\]
\end{parts}
\question Give a combinatorial proof for the identity $P(n,k) = {n \choose k}k!$
\end{questions}


\end{document}


