\documentclass[11pt]{exam}

\usepackage{amsmath, amssymb, multicol}
\usepackage{graphicx}
\usepackage{textcomp}


\def\d{\displaystyle}
\def\b{\mathbf}
\def\R{\mathbf{R}}
\def\Z{\mathbf{Z}}
\def\st{~:~}
\def\bar{\overline}
\def\inv{^{-1}}
\def\pow{\mathcal P}


%\pointname{pts}
\pointsinmargin
\marginpointname{pts}
\addpoints
\pagestyle{head}
%\printanswers

\firstpageheader{Math 228}{\bf Animal Parades for Kids}{Wednesday, Feb 11}


\begin{document}

%space for name
%\noindent {\large\bf Name:} \underline{\hspace{2.5in}}
%\vskip 1em

\begin{questions}
\question Suppose you have a huge box of animal crackers containing a plenty of each of 10 different animals.  For the counting questions below, carefully examine their similarities and differences, and then give an answer.  The answers are all one of the following.

\begin{choices}
\begin{multicols}{4}
\choice $ P(10,6)$
\choice ${10 \choose 6}$
\choice $10^6$
\choice ${15 \choose 9}$
\end{multicols}
\end{choices}

\begin{parts}
\part How many animal parades can you line up containing 6 crackers?
\vfill
\part How many animal parades of 6 crackers can you line up so that the animals appear in alphabetical order?
\vfill
\part How many ways could you line up 6 different animals in alphabetical order?
\vfill
\part How many ways could you line up 6 different animals if they can come in any order?
\vfill
\part How many ways could you give 6 children one animal cracker each?
\vfill
\part How many ways could you give 6 children one animal cracker each so that no two kids get the same animal?
\vfill
\part How many ways could you give out 6 giraffes to 10 kids?
\vfill 
\part Write a question about giving animal crackers to kids that has the answer ${10\choose 6}$.
\end{parts}
\end{questions}

\end{document}


